### 2018-05-14 1_intro.tex
write interpreter. update pictures.
cleanup a bit.

Most modern computer systems contain large parts that operate concurrently.
Although the parallelisation of a system can drastically improve its performance, it opens numerous of problems connected to correctness, robustness and reliability, which makes the concurrent program design one of the most difficult problems of programming~\cite{mckenney2017parallel}.

@@

As an example, consider a shared memory concurrent system that executes two parallel processes as it is described in Figure~\ref{simple_wmm_x86}.
The process~\texttt{P0} writes the value~\texttt{1} to the shared variable~\texttt{x} and reads a value of the shared variable~\texttt{y}, and the process~\texttt{P1} writes the~\texttt{y} and reads the~\texttt{x}.
Considering all interleavings of the processes, one could expect the final state to be one of the following states:
\begin{itemize}[noitemsep,topsep=0pt]
\item \texttt{(0:EAX=0,\,1:EBX=1)}
\item \texttt{(0:EAX=1,\,1:EBX=0)}
\item \texttt{(0:EAX=1,\,1:EBX=1)}
\end{itemize}
%As the program specifies the write to~\texttt{x} to precede the read of~\texttt{y} in process~\texttt{P0}, one could expect the final state `\texttt{(0:EAX=0~/\textbackslash~1:EAX=0)}' to be unreachable.
However, on the x86 architecture the state `\texttt{(0:EAX=0,\,1:EAX=0)}' is also reachable as the processors may cache the write to shared memory into their local write buffers, so that they do not immediately become visible by processes running on other cores.
This behaviour is known as \textit{store buffering~(SB)}.

@@

Such little examples that illustrate specific behaviour of a concurrent execution environment are called \textit{litmus tests}.
The formal way to define the semantics of memory operations and synchronisation primitives of a parallel execution environment is to define its \textit{memory model}.
%The behaviour of a concurrent execution environment is described by a formal memory model
%The study of concurrency of a specific platform concerns on defining
%and are widely used to illustrate its behaviour.
%Properties of an execution environment can be described by a formal \textit{memory model}
%The issues like one discussed above have led to the need for formalisation of semantics of memory operations within different concurrent architectures defined by \textit{memory models}.
There are two types of formal memory models. Models of the first type characterise the behaviour of the system (its \textit{operational semantics}) in terms of the abstract machine executing the code, as it was done for the SB example above.
Models of the second type define the \textit{axiomatic semantics} of the system by specifying the set of assertions over states of the program.
%There are two ways to define a memory model: by specifying its \textit{operational semantics} (in terms of an abstract machine executing the code, as it was done for the SB example above) or by defining its \textit{axiomatic semantics} (in terms of assertions over states of the program formed by sequences of operations).
Although the former type of memory models may be easier to describe and interpret, existing numerous formal verification tools and methods direct the research towards axiomatic memory models.

The first memory model for concurrent systems was formulated by Leslie Lamport back in 1979~\cite{lamport1979make}.
This memory model, called the \textit{sequential consistency (SC)}, allows only those executions that produce the same result as if the operations had been executed in an interleaved fashion in a single process%
%
\footnote{In order to prescind from the implementation details while discussing the memory models theory, we avoid the use of the software-specific term \textit{thread} and the hardware-specific term \textit{processor}. Instead, we adhere the terminology of the theory of concurrency employed by Ben-Ari~\cite{ben2006principles} by naming a concurrent piece of code the \textit{process}.}. %
%
This means that the order of operations executed by a process is strictly defined by the program (the code) it executes.
The SC model does require the write to a shared variable performed in one process to become visible by all other processes \textit{instantly}, each process writes directly to the shared memory, without local buffering.
Another important requirement of the SC memory model is that it forbids memory operations reordering within a single process (the order is strictly defined by the program).
Originally, the operational semantics was defined for the SC model, however there exist axiomatic specifications for it~\cite{mansky2015axiomatic}.

The SC model is considered to be a \textit{strong memory model} in the sense that it provides firm guarantees regarding the ordering and caused effect of memory operations.
Different relaxations of this model lead to the class of \textit{weak memory models~(WMMs)} that specify how processes interact through shared memory, when a write becomes visible to other processes and what value a read can return.
Thus, WMMs serve as set of guarantees made by designers of execution environment (hardware, programming language, compiler, database, operation system, etc.) to programmers on which behaviours of their concurrent code they may rely on.

@@

Research on weak memory models firstly aims to \textit{formalise} a formal approach of understanding programs with respect to weak memory models which is \textit{systematic}, \textit{sound} and \textit{complete}.
One of the most well-known frameworks for weak memory model-aware analysis was formalised by J. Alglave in~2010~\cite{alglave2010shared},.
In addition to developing rather theoretical basis, researchers work on extracting the WMMs for hardware architectures from existing implementations of from their specifications, which are written in natural language and thus suffer from ambiguities and incompleteness.
Over the last decade the memory models have been defined for most mainstream multiprocessor architectures, such as x86-TSO and Sparc-TSO (for \textit{Total Store Order}) model for x86 and Sparc architectures~\cite{owens2009better}, much more relaxed memory model for Power and ARM architectures~\cite{alglave2009semantics,sarkar2011understanding, alglave2014herding}, etc. %, or Alpha architecture~\cite{alphaWMM}~, and others.
There are projects for validating hardware architectures wrt. a memory model as well, e.g.~\cite{lustig2014pipecheck,lustig2016coatcheck}.

@@

The current work aims to rework the proof-of-concept tool \textit{\porthos{}} by extending the input language, which currently represents the minimum subset of C, and revising the general architecture of the tool in order to enhance performance, extensibility, reliability and maintainability.
%interprocedural analysis
As the general architecture and almost all components of \porthos{} have been redesigned, the tool received a new name -- \textit{\porthos[2]}\,%
%
\footnote{Hereinafter with the name `\porthos{}' we refer to the tool \porthos{} of version 1, whereas the new implementation of \porthos{} is called \porthos[2].}. %
%
Considering the enhancements of the architectural design, \porthos[2] represents a generalised framework for SMT-based memory model-aware analysis, which can not only perform the portability analysis, but can serve as a basis for other kinds of static code analysis.

>>>>

Most modern computer systems contain large parts that operate concurrently.
Although the parallelisation of a system can drastically improve its performance, it opens numerous of problems regarding correctness, robustness and reliability, which makes the concurrent program design one of the most difficult problems of programming~\cite{mckenney2017parallel}.

@@

As an example, consider a shared memory concurrent system that executes two parallel processes as it is described in Figure~\ref{simple_wmm_x86} (such little examples that illustrate specific behaviour of a concurrent execution environment are called \textit{litmus tests}).
The process~\texttt{P0} writes the value~\texttt{1} to the shared variable~\texttt{x} and reads a value of the shared variable~\texttt{y}, and the process~\texttt{P1} writes the~\texttt{y} and reads the~\texttt{x}.
Considering all interleavings of the processes, one could expect the final state to be one of the following:
\begin{itemize}[noitemsep,topsep=0pt]
\item \texttt{(0:EAX=0,\,1:EBX=1)},
\item \texttt{(0:EAX=1,\,1:EBX=0)},
\item \texttt{(0:EAX=1,\,1:EBX=1)}.
\end{itemize}
%As the program specifies the write to~\texttt{x} to precede the read of~\texttt{y} in process~\texttt{P0}, one could expect the final state `\texttt{(0:EAX=0~/\textbackslash~1:EAX=0)}' to be unreachable.
However, on the x86 architecture the state `\texttt{(0:EAX=0,\,1:EAX=0)}' is also reachable as the processors may cache the write to shared memory into their local write buffers, so that the write does not immediately become visible by processes running on other cores.
This behaviour is known as \textit{store buffering~(SB)}.

@@

The formal way to define the semantics of memory operations and synchronisation primitives of a parallel execution environment is to define its \textit{memory model}.
%The behaviour of a concurrent execution environment is described by a formal memory model
%The study of concurrency of a specific platform concerns on defining
%and are widely used to illustrate its behaviour.
%Properties of an execution environment can be described by a formal \textit{memory model}
%The issues like one discussed above have led to the need for formalisation of semantics of memory operations within different concurrent architectures defined by \textit{memory models}.
There are two main types of formal memory models.
Models of the first type characterise the behaviour of the system (its \textit{operational semantics}) in terms of the abstract machine executing the code, as it was done for the SB example above.
Models of the second type define the \textit{axiomatic semantics} of the system by specifying the set of assertions over states of the program.
%There are two ways to define a memory model: by specifying its \textit{operational semantics} (in terms of an abstract machine executing the code, as it was done for the SB example above) or by defining its \textit{axiomatic semantics} (in terms of assertions over states of the program formed by sequences of operations).
Although the former type of memory models may be easier to describe and interpret, existing numerous formal verification tools and methods address the research towards axiomatic memory models.

The first memory model for a concurrent system was formulated by Leslie Lamport back in 1979~\cite{lamport1979make}.
This memory model, called the \textit{sequential consistency (SC)}, allows only those executions that produce the same result as if the operations had been executed in an interleaved fashion in a single process%
%
\footnote{In order to prescind from the implementation details while discussing the memory models theory, we avoid the use of the software-specific term \textit{thread} and the hardware-specific term \textit{processor}. Instead, we adhere the terminology of the theory of concurrency employed by Ben-Ari~\cite{ben2006principles} by naming a concurrent piece of code the \textit{process}.}. %
%
This means that the order of operations executed by a process is strictly defined by the program (the code) it executes.
The SC model does require the write to a shared variable performed in one process to become visible by all other processes \textit{instantly} as each process writes directly to the shared memory, without local buffering.
Another important requirement of the SC memory model is that it forbids reordering of memory operations within a single process (the order is strictly defined by the program).
Originally, the operational semantics was defined for the SC model, however there exist axiomatic specifications for it~\cite{mansky2015axiomatic}.

The SC model is considered to be a \textit{strong memory model} in the sense that it provides firm guarantees regarding the ordering and effect of memory operations.
Different relaxations of the SC model lead to \textit{weak memory models~(WMMs)} that specify how processes interact through the shared memory, when a write becomes visible to processes running on other cores, and what value a read operation can get.
Thus, WMMs serve as set of guarantees made by designers of execution environment (hardware, programming language, compiler, database, operation system, etc.) to programmers on which behaviours of their concurrent code they can rely on.

@@

Research on weak memory models firstly aims to \textit{formalise} a formal approach of understanding programs with respect to weak memory models, which is \textit{systematic}, \textit{sound} and \textit{complete}.
One of the most well-known frameworks for weak memory model-aware analysis was formalised by J. Alglave in~2010~\cite{alglave2010shared}.
It is the event-based non-deterministic model without global time (see Section~\ref{ch:wmm:event} for details).

In addition to developing the theoretical basis, researchers work on extracting the memory models for hardware architectures from existing implementations or from the specifications, which are written in natural language and thus suffer from ambiguities and incompleteness.
Over the last decade the memory models have been defined for most mainstream multiprocessor architectures, such as x86-TSO and Sparc-TSO (for \textit{Total Store Order}) model for x86 and Sparc architectures~\cite{owens2009better}, much more relaxed memory model for Power and ARM architectures~\cite{alglave2009semantics,sarkar2011understanding, alglave2014herding}, etc. %, or Alpha architecture~\cite{alphaWMM}~, and others.
There are projects for validating hardware architectures wrt. a memory model as well, e.g.~\cite{lustig2014pipecheck,lustig2016coatcheck}.

@@

The current work aims to rework the proof-of-concept tool \textit{\porthos{}} by extending the input language, which currently represents the minimum subset of C, and revising the general architecture of the tool in order to enhance performance, extensibility, reliability and maintainability.
One of the directions of development was the ability to process the \textit{kernel litmus tests} written in C~\cite{TODO: KERNEL LITMUS TESTS WORK+REPOSITORY}.
%interprocedural analysis
As the general architecture and almost all components of \porthos{} have been redesigned, the tool received a new name \textit{\porthos[2]}\,%
%
\footnote{Hereinafter with the name `\porthos{}' we refer to the tool \porthos{} of version 1, whereas the new implementation of \porthos{} is called \porthos[2].}. %
%
Considering the enhancements of the architectural design, \porthos[2] represents a generalised framework for SMT-based memory model-aware analysis, which can not only perform the portability analysis, but can serve as a basis for other kinds of static code analysis.

### 2018-05-14 2_wmm.tex
cleanup a bit.

A \textit{computation event} $e_c \in \mathbb{C} \subseteq \mathbb{E}$, represents a low-level assembly computation operation performed solely on local-memory arguments.
An example of computation event may be the event $e_c = r_1 \leftarrow add(r_2, 1)$ that writes the sum of values stored in register $r_2$ and constant $1$ (which is modelled as a register as well) to the register $r_1$.
For modelling branching statements, we distinguish the set $\mathbb{C_{g}}~\subseteq~\mathbb{C}$ of \textit{guard} computation events (also called as \textit{branching events}), that are evaluated as a boolean value.

>>>>

A \textit{computation event} $e_c \in \mathbb{C} \subseteq \mathbb{E}$, represents a low-level assembly computation operation performed solely on local-memory arguments.
An example of computation event may be the event $e_c = r_1 \leftarrow add(r_2, 1)$ that writes the sum of values stored in register $r_2$ and constant $1$ (which is modelled as a register as well) to the register $r_1$.
For modelling branching statements, we define the set $\mathbb{C}_{g}~\subseteq~\mathbb{C}$ of \textit{guard} computation events (also called as \textit{branching events}), that are evaluated to a boolean value.

### 2018-05-14 4_impl.tex
cleanup a bit.

Considering the new meaning of the unrolling bound, we create sink events of two types: the \textit{complete} and \textit{incomplete} sink event.
The first one means that the loop has been executed a whole number of times, and the latter means that the unrolling bound triggered on not-the-last event of the loop.
Figure~\ref{fig:loop-unroll} illustrates unrolling for the cyclic control-flow graph on the left-hand side (the square node $S_{+}$ denotes the complete sink node, while the $S_{-}$ -- incomplete).

>>>>

Considering the new meaning of the unrolling bound, we create sink events of two types: the \textit{complete} and \textit{incomplete} sink event.
The first one means that the loop has been executed a whole number of times, and the latter means that the unrolling bound triggered on not-the-last event of the loop.
Figure~\ref{fig:loop-unroll} illustrates unrolling for the cyclic control-flow graph on the left-hand side (the square node $S_{+}$ denotes the complete sink node, and the $S_{-}$ denotes the incomplete sink).
